Regularised parameter selection in the non-stationary Hawkes process

### 2017-04-24 _paper_output/paper_sparse_hawkes.tex

We are interested in the problem of simultaneous identification of the
parameters of a Hawkes point process and identifying a sparse driving
signal by means of $\ell_1$ penalised regression where, by assumption,
the driving noise has a different form. This is analaogous to a blind
deconvolution problem or a compressive sensing-style inverse problem,
but considerably more challenging; While the process model superfcially
resembles a linear system, there are few of the conveneient results that
woudl apply in a linear regression problem; and as a time series problem
it is difficult to choose our penalty parameters by standard
non-parametric means such as cross validation. The proposed method uses
a heuristically-motivated Akaike-style Information criterion
regularisation parameter selection, which we test through simulation.
The proposed method works well in practice.

@@

We define call a counting process $N$ indexed by $t$ a (TODO)
predictable Poisson point process if it has an associated
\emph{intensity} process $\lambda$, such that $\lambda$ is predictable
with respect to the filtration generated by $N$, i.e.
\[\lambda(\mathcal{F(t^-)}):=\lim_{h\to 0} \frac{\mathrm E\left(N(t,t+h)\right)}{h}.\]

The Hawkes process \cite{hawkes_point_1971} specifically, is the linear,
self-exciting point process; that is, given a "background" rate
$\mu:\mathbb{R}^+\rightarrow\mathbb{R}^+,$ and a non-negative influence
kernel with positive support
$\phi:\mathbb{R}^+\rightarrow\mathbb{R}^+\cap\{0\}$, the Hawkes process
has the following expression for its intensity:

@@

In this problem, we are given a signal realisation of the count process
$N$ on an interval $T=[t_0,t_N]$. We are interested in identifying, from
this data, two unknown functional parameters; both the background rate
function $\mu$ and the influence kernel $\phi$.

    One such assumption would be penalization -{}-I apply an additional term
to the log likelihood function, penalizing particular values of the
parameters. For the penalized log likelihood for parameter $\theta$ and
penalty $\pi$, I write

@@

For non-parametric extension to a parametric model, one usually
penalizes only the non-parametric extensions to the model, such that
they vanish as the penalty term increases. :citeGreen1987

@@

For this particular model I have found few results giving me analytic
guarantees, so I use heuristics and simulation-based verification. I do
not declare that this method "best" in any sense, or that it will find
the correct global maximum penalized likelihood etc. The method,
however, is simple enough to prototype rapidly and, as it turns out,
performs well on test data. Results will be presented in the next
chapter.

@@

Similarly, we will approximate $\phi$ in a different functional basis
${f}_{1\leq k \leq K}$, such that $\phi(s):= \sum_{k=1}^K f(s;k)$. This
choice is likely arbitrary, but many exist in the literature trading off
computational and statistical convenience, including Laguerre
polynomials \cite{ogata_linear_1982}, Exponential functions
\cite{schoenberg_conistent_2005,rambaldi_modeling_2015}, and step
functions \cite{eichler_graphical_2016} As a convenient choice, we
propose a basis of smooth unimodal non-negative functions with a
computationally convenient form and a single mode, the convenient
Maxwell distribution,
\[\phi_j(t):=\sqrt{\frac{2}{\pi}} \frac{t^2\exp\left(-\frac{t^2}{2\sigma_j^2}\right)}{\sigma_j^3}.\]

@@

    \subsubsection{Optimization}\label{optimization}

The non-penalised log-likelihood function for a given realisation of the
data is given

@@

And using the method of maximum likelihood we estimate our parameter
vector by numerical optimisation of
\[\hat{\theta}_\pi(\mathbf t) = \operatorname{argmax}_\theta L_\pi(\mathbf t;\theta).\]

There is no closed-form solution for the optimum, but the derivatives of
this function are easy to calculate manually see e.g.
\cite{ozaki_maximum_1979}. Here howevfer, we use automatic
differentiation to make more complex penalisation procedure conveneient
to differentiate \cite{baydin_automatic_2014}.

We are able to use these to devise rapid gradient-pased optimisation
procedure.

@@

    \begin{center}
    \adjustimage{max size={0.9\linewidth}{0.9\paperheight}}{paper_sparse_hawkes_files/paper_sparse_hawkes_13_0.pdf}
    \end{center}
    { \hspace*{\fill} \\}

>>>>

We are interested in the problem of simultaneous identification of the
parameters of a Hawkes point process and identifying a sparse driving
signal by means of $\ell_1$ penalised regression where, by assumption,
the driving noise has a different form. This is analaogous to a blind
deconvolution problem or a compressive sensing-style inverse problem,
but considerably more challenging; While the process model superfcially
resembles a linear system, there are few of the conveneient results that
would apply in a linear regression problem; and as a time series problem
it is difficult to choose our penalty parameters by standard
non-parametric means such as cross validation. The proposed method uses
a heuristically-motivated degrees-of-freedom regularisation parameter
selection, which we test through simulation. The proposed method works
well in practice.

@@

We define call a counting process $N$ indexed by $t$ a (TODO)
predictable Poisson point process if it has an associated
\emph{intensity} process $\lambda$, such that $\lambda$ is predictable
with respect to the filtration generated by $N$, i.e.

\[\lambda(\mathcal{F(t^-)}):=\lim_{h\to 0} \frac{\mathrm E\left(N(t,t+h)\right)}{h}.\]

The Hawkes process \cite{hawkes_point_1971} specifically, is the linear,
self-exciting point process; Concretely, given a "background" rate
$\mu:\mathbb{R}^+\rightarrow\mathbb{R}^+,$ and a non-negative influence
kernel with positive support
$\phi:\mathbb{R}^+\rightarrow\mathbb{R}^+\cap\{0\}$, the Hawkes process
has the following expression for its intensity:

@@

In this problem, we are given a single realisation of the count process
$N$ on an interval $T=[t_0,t_N]$. We are interested in identifying, from
this data, two unknown functional parameters; both the background rate
function $\mu$ and the influence kernel $\phi$. Since this is a general
nonparametric problem we will use traditional regularisation techniques
to control the expolosion of parameters.

    We hope to recover the excitation kernel, which is unknown, and also a
background signal, by appropriate penalisation of the terms.

For the penalized log likelihood for parameter $\theta$ and penalty
$\pi$, I write

@@

For non-parametric extension to a parametric model, one usually
penalizes only the non-parametric extensions to the model, such that
they vanish as the penalty term increases. \cite{green_penalized_1987}

@@

@@

Similarly, we will approximate $\phi$ in a different functional basis
${f}_{1\leq k \leq K}$, such that $\phi(s):= \sum_{k=1}^K f(s;k)$. This
choice is likely arbitrary, but many exist in the literature trading off
computational and statistical convenience, including Laguerre
polynomials \cite{ogata_linear_1982}, Exponential functions
\cite{schoenberg_consistent_2005,rambaldi_modeling_2015}, and step
functions \cite{eichler_graphical_2016} As a convenient choice, we
propose a basis of smooth unimodal non-negative functions with a
computationally convenient form and a single mode, the convenient
Maxwell distribution,
\[\phi_j(t):=\sqrt{\frac{2}{\pi}} \frac{t^2\exp\left(-\frac{t^2}{2\sigma_j^2}\right)}{\sigma_j^3}.\]

@@

    The non-penalised log-likelihood function for a given realisation of the
data is given \cite{ozaki_etimating_1979}

@@

Using the method of maximum likelihood we estimate our parameter vector
by numerical optimisation of
\[\hat{\theta}_\pi(\mathbf t) = \operatorname{argmax}_\theta L_\pi(\mathbf t;\theta).\]

    There is no closed-form solution for the optimum of the Hawkes model
likelihood, penalised or not. The derivatives of the unpenalised
likelihood are easy to calculate manually -{}-{}- see e.g.
\cite{ozaki_maximum_1979}. Here, however, we use automatic
differentiation to make more complex penalisation procedure convenient
to differentiate \cite{baydin_automatic_2014}. The optimisation proceeds
using a standard black-box based Truncated Newton algorithm; as more
specialised sparse regression algorithms such as Least-Angle Regression,
do not directly apply here.

@@

    \begin{center}
    \adjustimage{max size={0.9\linewidth}{0.9\paperheight}}{paper_sparse_hawkes_files/paper_sparse_hawkes_14_0.pdf}
    \end{center}
    { \hspace*{\fill} \\}
